### 2021-07-08 rad-icfp-2021/appendix1.tex
.

\subsection{gradient from the introduction}
\label{sub:gradintro}

>>>>

\subsection{General arrays} % (fold)
\label{sub:lift_gen_arr}

We now show how to generalize our reverse-mode transformation to be defined on arrays over any ground type !G!.
That is, we need to adapt the reverse derivatives of !map2! and !reduce! when they have more general function arguments.

A ground type !G! is interpreted as an Euclidean space $A$. 
It is in particular a real vector space.
Similarly, a ground context !$\Gamma$=x$_1$:G$_1$,$\ldots$,x$_n$:G$_n$! is interpreted as $\bigoplus_{1\leq i\leq n}A_i$, where $A_i\defeq\seml$(!G$_i$!)$\semr$.
The denotation of the gradient of a term !$\Gamma \vdash$ e: G! at a point is then a matrix, more precisely an element of $(\bigoplus_{1\leq i\leq n}A_i)\otimes A$
where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}A_i\otimes A$.

We can define $\otimes$ on the types of our language inductively by

\begin{tabular}{r c l}
    $\reals \otimes A$ & $\defeq$ & $A$ \\
    $(A_1 \times \ldots \times A_n)\otimes A$ & $\defeq$ & $(A_1\otimes A) \times \ldots \times (A_n \otimes A)$ \\
    $\Array{A_1}{n} \otimes A$ & $\defeq$ & $\Array{A_1\otimes A}{n}$
\end{tabular}

With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A_1\times\ldots\times A_n$ as expected.
For !map2!, we need a generalization of $*:\reals\to\reals$. 
If !e$_1$: A!, then $\grad_{\{x\}}$!e$_1$: $A\otimes A$! and we need a new primitive $\widehat{*}:(A\otimes A) \times A \to A$.
If we represent $A\otimes A$ as a matrix, then $\widehat{*}$ is matrix-vector multiplication.
Similarly, for !A$_{3}$! in !reduce! we need a new primitive $\widetilde{*}:(A\otimes A)\times(A\otimes A) \to (A\otimes A)$.
$\widetilde{*}$ corresponds to matrix-matrix multiplication.
Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!, as can be seen in Figure~\ref{fig:lift_gen_arr}.

\begin{figure}
\begin{center}
\begin{tabular}{|r c l|}
\hline
    $\directD{\rho}{\Gamma}{Y}$(!map2 (x,y.e$_1$: G) e$_2$ e$_3$!) &=&  
    !let A,Y$_1$ = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
    && !let B,Y$_2$ = !$\directD{\rho}{\Gamma,A}{Y1}$!(e$_3$) in! \\
    && !let C=!$\grad_{\Gamma}$!e$_1$ in!\\
    && !<map2 (x,y.e$_1$) A B, fun (x$_1$,$\ldots$,x$_n$,Z) -> !\\
    && !Y$_2$( (x$_1$,$\ldots$,x$_n$)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x]!$\widehat{*}$!b) A Z,!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e$_1$)[a/x]!$\widehat{*}$!b) B Z )>!\\
    $\directD{\rho}{\Gamma}{Y}$(!reduce (x,y.e$_{1}$) e$_{2}$ e$_{3}$!) &=&
    !let y$_{1}$,Y$_{1}$ = !$\directD{\rho}{\Gamma}{Y}$!(e$_{2}$) in! \\
    && !let A,Y$_{2}$ = !$\directD{\rho}{\Gamma,y_1}{Y_1}$!(e$_{3}$) in! \\
    && !let A$_{0}$ = shift1R (scanl (x,y.e$_{1}$) y$_{1}$ A) in! \\
    && !let A$_{1}$ = shift1L (map2! \\ 
    && !      (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x,b/y]) A$_{0}$ A) in! \\
    && !let A$_{2}$ = map2! \\
    && !      (a,b.(!$\grad_{\{y\}}$!)e$_{1}$[a/x,b/y]) A$_{0}$ A in! \\
    && !let A$_{3}$ = scanr $\widetilde{*}$ 1 A$_{1}$ in! \\
    && !<reduce (x,y.e$_{1}$) y$_{1}$ A, fun (x$_{1}$,$\ldots$,x$_n$,z) ->! \\
    && !Y$_{2}$(x$_{1}$,$\ldots$,x$_n$, map2 (x,y. x$\widehat{*}$(y$\widehat{*}$z)) A$_{2}$ A$_{3}$>! \\ \hline
\end{tabular}
\end{center}
\caption{The reverse-mode AD transformation of \codekw{map2} and \codekw{reduce} for general arrays.}
\label{fig:lift_gen_arr}
\end{figure}

Evidently, one can combine all the generalizations from the previous subsections.
Even though this transformation has the correct complexity, it is open for future research to find
even better representations to allow for more optimizations. 
In particular, representations looking like Einsum \cite{van2011numpy} could be of interest 
and has been recently studied in the context of AD \cite{laue2018computing,laue2020simple}.
More generally, there is growing interest in tensor calculus \cite{liao2019differentiable,bernstein2020differentiating}.

\subsection{Gradient from the introduction}
\label{sub:gradintro}

### 2021-07-08 rad-icfp-2021/generalization.tex
.

\section{Beyond the Source language}
\label{sec:generalization}
In this section, we generalize our language by relaxing the imposed restrictions.
First, we allow free variables for the function argument of !reduce! in Section~\ref{sub:lift_reduce}. 
Then we show how to support more array operations (Section~\ref{sub:lift_more_arr_op}), conditionals (Section~\ref{sub:lift_conditional}), and non-smooth scalar operators (Section~\ref{sub:lift_non_smooth}).
Afterwards, we extend the language to support general arrays instead of supporting only $\Array{\reals}{n}$ in Section~\ref{sub:lift_gen_arr}. Finally, we provide a recipe for adding more constructs to the language in Section~\ref{sub:lift_recipe}.

@@

\subsection{General arrays} % (fold)
\label{sub:lift_gen_arr}

We now show how to generalize our reverse-mode transformation to be defined on arrays over any ground type !G!.
That is, we need to adapt the reverse derivatives of !map2! and !reduce! when they have more general function arguments.

A ground type !G! is interpreted as an Euclidean space $A$. 
It is in particular a real vector space.
Similarly, a ground context !$\Gamma$=x$_1$:G$_1$,$\ldots$,x$_n$:G$_n$! is interpreted as $\bigoplus_{1\leq i\leq n}A_i$, where $A_i\defeq\seml$(!G$_i$!)$\semr$.
The denotation of the gradient of a term !$\Gamma \vdash$ e: G! at a point is then a matrix, more precisely an element of $(\bigoplus_{1\leq i\leq n}A_i)\otimes A$
where $\otimes$ is the tensor product of real vector spaces. This space is isomorphic to $\bigoplus_{1\leq i\leq n}A_i\otimes A$.

We can define $\otimes$ on the types of our language inductively by

\begin{tabular}{r c l}
    $\reals \otimes A$ & $\defeq$ & $A$ \\
    $(A_1 \times \ldots \times A_n)\otimes A$ & $\defeq$ & $(A_1\otimes A) \times \ldots \times (A_n \otimes A)$ \\
    $\Array{A_1}{n} \otimes A$ & $\defeq$ & $\Array{A_1\otimes A}{n}$
\end{tabular}

With this definition, we recover that the gradient of $\Gamma \vdash$!e: !$\reals$ is a tuple of type $A_1\times\ldots\times A_n$ as expected.
For !map2!, we need a generalization of $*:\reals\to\reals$. 
If !e$_1$: A!, then $\grad_{\{x\}}$!e$_1$: $A\otimes A$! and we need a new primitive $\widehat{*}:(A\otimes A) \times A \to A$.
If we represent $A\otimes A$ as a matrix, then $\widehat{*}$ is matrix-vector multiplication.
Similarly, for !A$_{3}$! in !reduce! we need a new primitive $\widetilde{*}:(A\otimes A)\times(A\otimes A) \to (A\otimes A)$.
$\widetilde{*}$ corresponds to matrix-matrix multiplication.
Then, using this notation, there are only minimal changes to the reverse derivatives of !map2! and !reduce!, as can be seen in Figure~\ref{fig:lift_gen_arr}.

\begin{figure}
\begin{center}
\begin{tabular}{|r c l|}
\hline
    $\directD{\rho}{\Gamma}{Y}$(!map2 (x,y.e$_1$: G) e$_2$ e$_3$!) &=&  
    !let A,Y$_1$ = !$\directD{\rho}{\Gamma}{Y}$!(e2) in! \\
    && !let B,Y$_2$ = !$\directD{\rho}{\Gamma,A}{Y1}$!(e$_3$) in! \\
    && !let C=!$\grad_{\Gamma}$!e$_1$ in!\\
    && !<map2 (x,y.e$_1$) A B, fun (x$_1$,$\ldots$,x$_n$,Z) -> !\\
    && !Y$_2$( (x$_1$,$\ldots$,x$_n$)!$\widehat{+}$!(C!$\widehat{*}$!(reduce + 0 Z)),!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{x\}}$!e$_1$)[a/x]!$\widehat{*}$!b) A Z,!\\
    && \quad\quad! map2 (a,b.(!$\grad_{\{y\}}$!e$_1$)[a/x]!$\widehat{*}$!b) B Z )>!\\
    $\directD{\rho}{\Gamma}{Y}$(!reduce (x,y.e$_{1}$) e$_{2}$ e$_{3}$!) &=&
    !let y$_{1}$,Y$_{1}$ = !$\directD{\rho}{\Gamma}{Y}$!(e$_{2}$) in! \\
    && !let A,Y$_{2}$ = !$\directD{\rho}{\Gamma,y_1}{Y_1}$!(e$_{3}$) in! \\
    && !let A$_{0}$ = shift1R (scanl (x,y.e$_{1}$) y$_{1}$ A) in! \\
    && !let A$_{1}$ = shift1L (map2! \\ 
    && !      (a,b.(!$\grad_{\{x\}}$!e$_{1}$)[a/x,b/y]) A$_{0}$ A) in! \\
    && !let A$_{2}$ = map2! \\
    && !      (a,b.(!$\grad_{\{y\}}$!)e$_{1}$[a/x,b/y]) A$_{0}$ A in! \\
    && !let A$_{3}$ = scanr $\widetilde{*}$ 1 A$_{1}$ in! \\
    && !<reduce (x,y.e$_{1}$) y$_{1}$ A, fun (x$_{1}$,$\ldots$,x$_n$,z) ->! \\
    && !Y$_{2}$(x$_{1}$,$\ldots$,x$_n$, map2 (x,y. x$\widehat{*}$(y$\widehat{*}$z)) A$_{2}$ A$_{3}$>! \\ \hline
\end{tabular}
\end{center}
\caption{The reverse-mode AD transformation of \codekw{map2} and \codekw{reduce} for general arrays.}
\label{fig:lift_gen_arr}
\end{figure}

Evidently, one can combine all the generalizations from the previous subsections.
Even though this transformation has the correct complexity, it is open for future research to find
even better representations to allow for more optimizations. 
In particular, representations looking like Einsum \cite{van2011numpy} could be of interest 
and has been recently studied in the context of AD \cite{laue2018computing,laue2020simple}.
More generally, there is growing interest in tensor calculus \cite{liao2019differentiable,bernstein2020differentiating}.

>>>>

\section{Beyond the Source language}
\label{sec:generalization}
In this section, we generalize our language by relaxing the imposed restrictions.
First, we allow free variables for the function argument of !reduce! in Section~\ref{sub:lift_reduce}. 
Then we show how to support conditionals in Section~\ref{sub:lift_conditional}.
% Then we show how to support more array operations (Section~\ref{sub:lift_more_arr_op}), conditionals (Section~\ref{sub:lift_conditional}), and non-smooth scalar operators (Section~\ref{sub:lift_non_smooth}).
% Afterwards, we extend the language to support general arrays instead of supporting only $\Array{\reals}{n}$ in Section~\ref{sub:lift_gen_arr}. 
Finally, we provide a recipe for adding more constructs to the language in Section~\ref{sub:lift_recipe}. Further generalizations for more array operations, non-smooth scalar operators, and general array support can be found in the supplementary materials.

@@
